Automatic Code Cleaning in R with `Rclean`

### 2020-03-13 ropensci/rclean_technote.Rmd
Typo fix
Edits throughout for final version

@@

Data analysis can be messy and complicated. Because of project
complexities and time contraints, analytial software often reflects
these challenges. "What did I measure? What analyses are relevant to
them? Do I need to transform the data? What's the function for the
analysis I want to run?" In addition, although many researchers see
the value in learning to write software, the entry level is still
exceedingly high. The `R` language has become very popular among
scientists and analysts because of the lower threshold, which is good
in that it empowers science. The downside is that learning to create
good software is a skill that takes a considerable amount of time to
learn, which means that many scientists write code that will get the
job "done" but lacks transparency and is highly unstable (i.e. buggy).

The goal of `Rclean` is to provide a tool that help someone reduce and
organize code based on results.  More often then not, when someone is
writing an R script, the intent is to produce a set of results, such
as a statistical analysis, figure or table. This set of results is
nearly always a subset of a much larger set of possible ways to
explore a dataset, as there are many statistical approaches and tests,
let alone ways to create visualizations and other representations of
patterns in data. This commonly leads to lengthy, complicated scripts
from which researchers manually subset results, but likely never to be
refactored because of the difficulty in disentangling the code needed
to produce some results and not others.

The `Rclean` package uses an automated technique based on data
provenance and network algorithms to analyze existing scripts and
provide ways to identify and extract code to produce a desired output
without having to manually conduct the search through the code line by
line. `Rclean` was designed to ease refactoring for scientists that
use R but do not have formal training in software engineering and
specifically with the "art" of creating clean code; however, because
manually culling code is tedious and potentially leads to errors,
`Rclean` can be a useful tool for everyone at all levels of coding
ability.

There are multiple approaches to collecting data provenance, but
`Rclean` uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. `Rclean` relies on a library called
`CodeDepends` to gather the prospective provenance for each
script. For more information on the mechanics of the `CodeDepends`
package, see [@R-CodeDepends]. To get an idea of what data provenance
is, take a look at the `code_graph` function. The plot that it
generates is a graphical representation of the prospective provenance
generated for `Rclean` \ref{fig:prov-graph}.

The provenance provides a network representation of relationships
among functions and objects. This is very powerful because we can now
apply algorithms analyze the R script with respect to our results. The
main function of `Rclean`, called `clean`, takes the provenance and
applies a depth first network search algorithm to determine the
pathways leading from inputs to outputs, excluding any objects or
lines of code that do not fall along that pathway and are, therefore,
not necessary to produce the desired set of results. As demonstrated
in the next section, through this process, we can produce a new script
with the minimal code necessary to generate the output we want.

# Installation

@@

# Using `Rclean` to clean your code

@@

So, let's say we've come to our script wanting to extract the code to
produce one of the results `fit_sqrt_A`, which is an analysis that is
the fitted model object for an analysis. We might want to double check
the results, and we also might need to use the code again for another
purpose, such as creating a plot of the patterns supported by the
test. Manually tracing through our code for all the variables used in
the test and finding all of the lines that were used to prepare them
for the analysis would be annoying and difficult, especially given the
fact that we have used "x" as a prefix for multiple unrelated objects
in the script. However, `Rclean` can do this easily via the `clean`.

@@

Especially when the code for different variables are entangled, it can
be useful to visual the code in order to devise an approach to
cleaning. As seen above \ref{fig:prov-graph}, `code_graph` will
produce a visual of the code and the objects that they produce.

@@

@@

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring code. Using ``Rclean`` the
code necessary to produce a specified result (e.g., an object stored
in memory or a table or figure written to disk) can be easily and
*reliably* isolated even when tangled with code for other
results. Tools, such as this, that make it easier to produce
transparent, accessible code will be an important aid for improving
scientific reproducibility.

Relatedly, it is important to point out that `Rclean` *does not* keep
comments present in code. This is a limitation of the data provenance,
which currently does not assign them a relationships. Therefore,
although there is often very useful or even invaluable information in
comments, the `clean` function removes them. This is a general issue
with automated methods for detecting the relationships between
comments and the code itself. Comments at the end of lines are
typically relevant to the line they are on, but this is not a
requirement and could refer to other lines. Also, comments occupying
their own lines usually refer to the following lines, but this is also
not necessarily the case. As `clean` depends on the unambiguous
determination of relationships in the production of results, it cannot
operate automatically on comments. However, comments in the original
code remain untouched and can be used to inform the reduced
code. Also, as the `clean` function is oriented toward isolating code
based on a specific result, the resulting code tends to naturally
support the generation of new comments that are higher level
(e.g. "The following produces a plot of the mean response of each
treatment group."), and lower level comments are not necessary because
the code is simpler and clearer. Lastly, as stated above, comments can
serve an important role in coding, it is worth reflecting on the
statement in R.C. Martin's book `Clean Code: A Handbook of Agile
Software Craftsmanship` that "Comments do not compensate for bad
code."

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as libraries that are actually used by the code,
processing comments, parsing control statements and replicating random
processes. Using retrospective provenance comes at a cost. In order to
gather it, the script needs to be executed. When scripts are
computationally intensive or contain bugs that stop execution, then
retrospective provenance can not be obtained for part or all of the
code. Some work has already been done in the direction of implementing
retrospective provenance for code cleaning in R (see
http://end-to-end-provenance.github.io). Although such costs may
present challenges, combining prospective and retrospective provenance
methods could provide a powerful and flexible solution.

Although the current implementation of Rclean for minimizing code is
useful on its own, we see promise in connecting it with other clean
code and reproducibility tools. One example is the `reprex` package,
which provides a simple API for sharing reproducible examples
[@R-reprex]. Another possibility is to help transition scripts to
function, package and workflow creation and refactoring via toolboxes
like `drake` [@R-drake]. `Rclean` could provide a reliable way to
extract parts of a larger script that would be piped to a simplified
reproducible example, in the case of `reprex`, or, since it can
isolate the code from inputs to one or more outputs, be used to
extract all of the components needed to write one or more functions
that would be a part of a package or workflow, as is the goal of
`drake`.

To conclude, we hope that `Rclean` makes writing scientific software
easier for the R community. As an open-source project, we look forward
to feedback and help with extending it, particularly in the area of
reproducibility. We invite people to use the package and get involved
by reporting bugs and suggesting or contributing features. For more
information please visit the project page on Github.

>>>>

# Introduction

@@

The `R` language has become very popular among scientists and analysts
because it enables the rapid development of software and empowers
scientific investigation. However, regardless of the language used,
data analysis is usually complicated. Because of various project
complexities and time contraints, analytical software often reflects
these challenges. "What did I measure? What analyses are relevant to
them? Do I need to transform the data? What's the function for the
analysis I want to run?" Although many researchers see the value in
learning to write software, the time investment for learning a
programming language alone is still exceedingly high for many, let
alone also learning software best practices. The downside to the rapid
spread of data science is that learning to create good software takes
a back-seat to just writing code that will get the job "done" leading
to issues with transparency and software that is highly unstable
(i.e. buggy and not reproducible).

The goal of `Rclean` is to provide a automated tool to help scientists
more easily write better code. Specifically, `Rclean` has been
designed to faciliate the isolation of the code needed to produce one
or more results, because more often then not, when someone is writing
an R script, the ultimate goal is analytical results for inference,
such as a set of statistical analyses and associated figures and
tables. As the investigative process is inherently iterative, this set
of results is nearly always a subset of a much larger set of possible
ways to explore a dataset. There are many statistical tests and
visualizations and other representations that can be employed in a
myriad of ways depending on how the data are processed. This commonly
leads to lengthy, complicated scripts from which researchers manually
subset results, but likely never to be refactored because of the
difficulty in disentangling the code.

The `Rclean` package uses a technique based on data provenance and
network algorithms to isolate code for a desired result
*automatically*. The intent is to ease refactoring for scientists that
use R but do not have formal training in software design and
specifically with the "art" of creating clean code, which in part is
the development of an intuitive sense of how code is and/or should be
organized. However, manually culling code is tedious and potentially
leads to errors, and we see `Rclean` as a useful tool for everyone at
all levels of coding ability in `R`. Here, we'll cover details on the
implementation and design of the package, a general example of how it
can be used and thoughts on its future development.

# How to Use `Rclean` to Write Cleaner Code

## Installation

@@

## Isolating Code for a Set of Results

@@

So, let's say we've come to our script wanting to extract the code to
produce one of the results `fit_sqrt_A`, which is an analysis that is
the fitted model object for an analysis. We might want to double check
the results, and we also might need to use the code again for another
purpose, such as creating a plot of the patterns supported by the
test. Manually tracing through our code for all the variables used in
the test and finding all of the lines that were used to prepare them
for the analysis would be annoying and difficult, especially given the
fact that we have used "x" as a prefix for multiple unrelated objects
in the script. However, `Rclean` can do this easily via the `clean`
function.

@@

Especially when the code for different variables are entangled, it can
be useful to visual the code in order to devise an approach to
cleaning. The `code_graph` function can also give us a visual of the
code and the objects that they produce.

@@

There are multiple approaches to collecting data provenance, but
`Rclean` uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. `Rclean` relies on a library called
`CodeDepends` to gather the prospective provenance for each
script. For more information on the mechanics of the `CodeDepends`
package, see [@R-CodeDepends]. To get an idea of what data provenance
is, we can use `code_graph` function to get a graphical representation
of the prospective provenance generated for `Rclean`
\ref{fig:prov-graph}.

All of this work with the provenance is to get the network
representation of relationships among functions and objects. The
provenance network is very powerful because we can now apply
algorithms analyze the R script with respect to our results. This is
what empowers the `clean` function, which takes the provenance and
applies a a network search algorithm to determine the pathways leading
from inputs to outputs. In the process any objects or lines of code
that do not fall along that pathway are by definition not necessary to
produce the desired set of results and can therefore be removed. As
demonstrated in the example, this property of the provenance network
is what facilitates the robust isolation of the minimal code necessary
to generate the output we want.

One important topic to discuss is that `Rclean` *does not* keep
comments present in code. This is the result of a limitation of the
data provenance collection, which currently does not assign them a
relationship in the provenance network. This is a general issue with
detecting the relationships between comments and code. For example,
comments at the end of lines are typically relevant to the line they
are on, but this is not a requirement and could refer to other
lines. Also, comments occupying their own lines usually refer to the
following lines, but this is also not necessarily the case either, as
there is no "linguistic" requirement for this. Although `Rclean`
cannot operate automatically on comments; however, comments in the
original code remain untouched and can be used to inform the reduced
code. Also, as the `clean` function is oriented toward isolating code
based on a specific result, the resulting code tends to naturally
support the generation of new comments that are higher level
(e.g. "The following produces a plot of the mean response of each
treatment group."), and lower level comments are not necessary because
the code is simpler and clearer. This process of comments is an
important part of writing better code. Lastly, although comments can
serve an important role in coding, it is worth reflecting on the
statement in R.C. Martin's book `Clean Code: A Handbook of Agile
Software Craftsmanship` where he writes that, "Comments do not
compensate for bad code."

@@

``Rclean`` provides a simple, easy to use tool for scientists who
would like help refactoring code. Using ``Rclean``, the code necessary
to produce a specified result (e.g., an object stored in memory or a
table or figure written to disk) can be easily and *reliably* isolated
even when tangled with code for other results. Tools, such as this,
that make it easier to produce transparent, accessible code will be an
important aid for improving scientific reproducibility
[@Pasquier2017].

Although the current implementation of Rclean for minimizing code is
useful on its own, we see promise in connecting it with other
reproducibility tools. One example is the `reprex` package, which
provides a simple API for sharing reproducible examples
[@R-reprex]. Another possibility is to help transition scripts to
function, package and workflow creation and refactoring via toolboxes
like `drake` [@R-drake]. `Rclean` could provide a reliable way to
extract parts of a larger script that would be piped to a simplified
reproducible example, in the case of `reprex`, or, since it can
isolate the code from inputs to one or more outputs, be used to
extract all of the components needed to write one or more functions
that would be a part of a package or workflow, as is the goal of
`drake`.

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as libraries that are actually used by the code,
processing comments (as discussed above), parsing control statements
and replicating random processes. Using retrospective provenance comes
at a cost, however. In order to gather it, the script needs to be
executed. When scripts are computationally intensive or contain bugs
that stop execution, then retrospective provenance can not be obtained
for part or all of the code. Some work has already been done in the
direction of implementing retrospective provenance for code cleaning
in R (see http://end-to-end-provenance.github.io). Although such costs
may present challenges, combining prospective and retrospective
provenance methods could provide a powerful and flexible solution.

We hope that `Rclean` makes writing scientific software easier for the
R community. We look forward to the future progress of the package and
other "code cleaning" tools. As an open-source project, we are very
receptive to feedback and help with extending it. We invite people to
use the package and get involved by reporting bugs and suggesting or
(hopefully) contributing features. For more information please visit
the project page on Github.
